Learning with misspecified models: Overconfidence and Stereotypes

### 2023-10-07 Paper/draft.tex
trap behavior

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Learning with misspecified models: Overconfidence and Stereotypes},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=2.5cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\title{Learning with misspecified models: Overconfidence and
Stereotypes}
\author{Jimena Galindo}
\date{September 07, 2023}

@@

\hypertarget{related-literature}{%
\section{Related Literature}\label{related-literature}}

\hypertarget{framework-1}{%
\section{Framework 1}\label{framework-1}}

An agent is of type \(\theta \in \Theta\) and faces an unknown exogenous
state \(\omega\) drawn from some density \(f\) over \(\Omega\). The
agent knows the distribution of \(\omega\) but not its realized value.
His prior belief about his type is \(p_0(\theta)\) and his belief about
the state state \(p_0(\omega)\) coincides with the true distribution
\(f\). Let the agent's type be \(\theta^*\) and the realized state be
\(\omega^*\).

An agent has a \emph{misspecified} belief if the prior assigns
probability zero to their true type. Furthermore, the agent is
\emph{dogmatic} if he holds a degenerate belief that places probability
one on being a particular type, \(\hat{\theta}\). An agent can be
dogmatic and misspecified; that means that
\(\hat{\theta} \neq \theta^*\) and \(p_0(\hat{\theta}) = 1\).

@@

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption is symplifies
the analysis and plays a role on whether an agent who updates their
beliefs using bayes rule would learn the truth or not.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
@Esponda2016 to this setting with only one agent}. A
\emph{self-defeating equilibrium} is a belief and action pair such that
the agent's belief about their type is misspecified and the outcome
generated by the action is consistent with the misspecified belief. The
average outcome under the true type and the true state equals the
average output the agent expects under the misspecified belief. In
addition, the agent's belief is said to be \emph{stable} when this
happens.

Within this framework, I consider two nested theories of belief
updating. The first one is a dogmatic modeler from \citet{Heidhues2018}.
The second one is a switcher, as in \citet{Ba2023}. The dogmatic modeler
can be seen as a switcher with an infinitely sticky initial belief and
this is the sense in which the two are nested. Both theories produce
different predictions with respect to the equilibrium outcome.

\hypertarget{the-dogmatic-modeler}{%
\subsection{The Dogmatic Modeler}\label{the-dogmatic-modeler}}

A dogmatic agent does not update their beliefs about \(\theta\); instead
he holds a degenerate belief that places probability one on being a
particular type, \(\hat{\theta}\), which is potentially not his true
type. In this case, no matter how much information he gathers against
being of type \(\hat{\theta}\), he will not update his beliefs. Any
discrepancies between the observed outcomes and his believed type are
incorporated using the Bayes rule to update their beliefs about
\(\omega\). This means the dogmatic agent will never learn their true
type if they startu off as being misspecified.

\citet{Heidhues2018} show that, under certain assumptions on the
per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{ee}<0$ and $u_e(\underline{e} \theta, \omega)>0>u_e(\Bar{e}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{e\theta}<0$ and $u_{e\omega}>0$. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes they observe reinforce their
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, they will be in a
self-defeating equilibrium whith a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome. He expected a
gain of \(h(\hat{\theta}, \mathbb{E}(\omega), a)\) but instead observes
\(h(\hat{\theta}, \omega^*, a)\). There are two possible sources for the
disappointment, the first is that the realized state is lower than the
expected state. The second source is that the agent is of type
\(\theta^*\) and therefore, for all possible states, his gain will be
lower than what he expected. Because the agent is dogmatic, he will not
update his beliefs about \(\theta\) and as a consequence will attribute
the disappointment to the state being lower than expected. He will
continue to update in ths way until he converges to a belief about
\(\omega\) that is stable. Such a belief will explain the observed
utility perfectly and allow the agent to rationalize his dogmatic belief
about \(\theta\). Under the assumptions of \citet{Heidhues2018}, there
is a unique value of \(\omega\) at which the belief is stable, I will
refer to such value as \(\omega_\infty\). This mechanism is further
illustrated in Example 1.

@@

\emph{Under this specification, the myopic optimal effort level is
\(a_t^* = \omega^*\). Nonetheless, because the agent does not know
\(\omega^*\), he will choose \(a_t = \mathbb{E}_t(\omega)\) where the
expectation is taken with respect to the agent's belief at the beginning
of period \(t\). If he does not revise his effort choice for \(k\)
periods, he will receive an average utility of
\((\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\) but he was expecting
an average utility of
\((\hat{\theta}\theta+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\). In
response, he will apply bayes rule to update his beliefs about
\(\omega\) to get the posterior belief with
\(\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}\)
which is lower than the initial belief. This will cause the agent to
choose a lower effort at \(t+k\). As a result, he will again receive an
average utlitity that is lower than what he expected which will cause
his belief to drift further down. This process will continue until the
average utility equals his expected utility under the dogmatic belief
that assigns probability 1 to \(\hat{\theta}\). At that point, the
stundent will have reached a self-defeating equilibrium and he will
continue to choose sub-optimal effort forever. }

Although the model of a dogmatic modeler is able to rationalize the
prevalence of overconfident (underconfident) beliefs, the assumption
that the agent has a degenerate belief and no mechanism through which he
can update such belief is very restrictive. An alternative approach is
proposed by \citet{Ba2023}. She proposes an extension of the dogmatic
agent who is able to jump from one dogmatic belief to another. By doing
so, the agent might end up being dogmatic and correctly speciefied.

\hypertarget{the-switcher}{%
\subsection{The Switcher}\label{the-switcher}}

An agent is a \emph{switcher} if they behave as a dogmatic, but is
willing to entertain the possibility that they are of a different type.
In particular, when they start off as a misspecified dogmatic, they are
willing to switch to a different dogmatic belief if the data is
convincing enough. Notice that their beliefs are still degenerate and
assign probability one to a particular type, and zero to all other
types. This means that a bayesian update on \(\theta\) would cause no
change in their beliefs. However, they are willing to entertain two such
beliefs and have a mechanism by which they decide which belief to adopt
at any period \(t\).

In order to abandon their initial dogmatic belief, the agent needs to
observe a sequence of outcomes that are sufficiently unlikely to have
happened if they were of the type they initially believed. They do so by
keeping track of the likelihood that each of the possible types
generated the data. If the likelihood ratio is sufficiently large, the
agent will switch to the alternative and behave as if they are dogmatic
about the new type.

In particular, for an agent that starts off with a dogmatic belief that
they are of type \(\hat{\theta}\) but is willing to consider the
alternative explanation that they are of type \(\tilde{\theta}\), the
agent will switch to the alternative if:

@@

Where \(h^t\) is the history of outcomes up to time \(t\) and \(\alpha\)
is the switching threshold. By keeping track of the likelihood ratio,
the agent can perform a \emph{Bayesian hypothesis test} and adopt the
Dogmatic belief that best fits the
data.\footnote{In a related problem @Schwarstein2021 proposes a similar updating procedure wich relies on the Bayesian 
hypothesis test.} Notice that if \(\alpha \to \infty\), the behavior of
the switcher will be indistinguishable from that of the Dogmatic
modeler. In this sense, the switcher is a generalization of the dogmatic
type.

By allowing the agent to keep track of the likelihoods and switching to
an alternative type, the switcher can avoid the self-confirming
equilibrium. However, if the prior belief on \(omega\) is sufficiently
tight around a self-defeating equilibrium, the switcher might look
identical to the dogmatic even in a case where \(\alpha\) is not too
large. This happens because under the agent's prior, the likelihood
ratio is unlikely to grow as fast as it is needed to escape the
self-defeating equilibrium. In this cases we say that the misspecified
belief is persistent. Hence, in order to be able to determine which of
these two models provides a better explanation of the observed
behaviors, we must focus on cases in which the prior is diffused enough
and the sequence of realized outcomes is such that the self-defeating
equilibrium would not persist.

\hypertarget{framework-2}{%
\section{Framework 2}\label{framework-2}}

The agent is still of some type \(\theta^* \in \Theta\) and the state is
\(\omega \sim F\). In this case the agent chooses an action \(a\in A\)
and observes a binary outcome that is either a success or a failure.
Denote the outcome by \(o \in {s,f}\). The probability of observing a
success is ingreasing in \(\theta^*\) and in \(\omega\). Whenever the
agent observes a succes, he gets a payoff \(v>0\) and whenever the
outcome is a failure, normalize the payoff to 0. In addition, the
probability of success is such that for each state, there is a unique
optimal action that maximizes the agent's expected payoff.

Two nested theories that have been widely studied fall within this
framework: Fully Bayesian updating and self-serving attribution bias. I
explain each of these classical models of belief updating in what
follows

\hypertarget{the-bayesian}{%
\subsection{The Bayesian}\label{the-bayesian}}

A Bayesian agent simultaneously updates their beliefs about \(\theta\)
and \(\omega\) by using Bayes' rule. The posterior odds at period t
about \(\theta\) after observing an outcome are given by: \[
\frac{p_{t}[\theta_H|\text{outcome}]}{p_{t}[\theta_M|\text{outcome}]} = 
      \frac{p[\text{outcome}|\theta_H]p_{t-1}[\theta_H]}{p[\text{outcome}|\theta_M]p_{t-1}[\theta_M]}
\] and \[
\frac{p_{t}[\theta_M|\text{outcome}]}{p_{t}[\theta_L|\text{outcome}]} = 
      \frac{p[\text{outcome}|\theta_M]p_{t-1}[\theta_M]}{p[\text{outcome}|\theta_L]p_{t-1}[\theta_L]}
\]

Where \(p_{t-1}\) is the prior at period \(t\) and
\(p[\text{outcome}|\theta] = \sum_{\omega} p[\text{outcome}|\theta, \omega, e]p_{t-1}(\omega)\)
is the probability of observing the outcome given the agent's type and
the effort chosen. The update is symmetric for \(\omega\).

Bayesian agents always choose the effort level that maximizes their flow
payoff by taking expectations over their prior beliefs about \(\theta\)
and \(\omega\). Since agents are myopic, even though all the parameters
could be identified with enough variation in choices, although the
Bayesian agent is the closest to a fully rational agent discussed here,
they might not learn their true type. This happens because, by being
myopic, they do not internalize the tradeoff between flow payoff and
learning. This can result in too little experimentation to learn their
true type. An alternative to this approach is given by
\citet{Hestermann2021} and is discussed with the results.

\hypertarget{the-self-serving-bayesian}{%
\subsection{The Self-Serving Bayesian}\label{the-self-serving-bayesian}}

A self-serving bayesian is an agent who uses a biased version of Bayes
rule to update his beliefs.He will update his beliefs about the state
\(\omega\) and his type \(\theta\) by over-attributing successes to a
high value of \(\theta\) and under-estimating the role of higher
\(\omega\). Similarly, he will attribute failure to a low state to a
greter degree than an unbiased agent would. To model the self-serving
attribution bias, I take the approach of \citet{benjamin2019}, where the
posterior is given by:

\[
p_{t}[\theta_H|\text{outcome}] = 
\frac{p[\text{outcome}|\theta]^{c_s^{\theta}\mathbb{I}\{\text{success}\}+c_f^{\theta}\mathbb{I}\{\text{failure}\}}p_{t-1}[\theta]}
{\sum_{\theta'\in\Theta}p[\text{outcome}|\theta']^{c_s^{\theta}\mathbb{I}\{\text{success}\}+c_f^{\theta}\mathbb{I}\{\text{failure}\}}p_{t-1}[\theta']}
\]

\(c_s^{\theta}\) and \(c_f^{\theta}\) are the self-serving attribution
bias parameters for the agent's type \(\theta\). If
\(c_s^{\theta} = c_f^{\theta} = 1\), the agent is unbiased and the
update is the same as the Bayesian update. On the other hand, if
\(c_s^{\theta} > c_f{\theta}\) the agent over-attributes success to
their type and under-attributes failure to their type\footnote{
  notice that the values of $c_s^{\theta}$ and $c_f^{\theta}$ are not restricted to be greater than 1. If they are both equal to 
  each other but less (more) than one, then the bias is simply underinference (overinference).}.

The update for \(\omega\) is analogous but with \(c_f^{\omega}\) and
\(c_s^{\omega}\) instead of \(c_f^{\theta}\) and \(c_s^{\theta}\) and
the bias is present whenever \(c_f^{\omega} > c_s^{\omega}\). That is,
the agent over-attributes failure to a low state relative the higher
states and under-attributes success to a low state relative to the
higher states.

@@

@@

By including self-confirming equilibria, the example captures the forces
from each of the updating mechanisms discussed in the previous section
and allows for direct comparison of all the theories. For realizations
of \((\theta, \omega)\) for which there are self-confirming equilibria,
the dogmatic agent will fall into the trap whereas the switcher will be
able to escape it. Similarly, an agent with self-attribution bias will
update their beliefs differently from an unbiased Bayesian, leading them
to choose different gambles. I exploit such cases in order to test which
model is a better fit for how subjects behave in a laboratory
experiment.
\footnote{because the setting does not match that of @Heidhues2018, there will be situationf for which the theory 
does not provie a prediction. If such cases arise in the lab, they will not be used for the analysis. However, whether a misspecified 
belief persists or not for the switcher, depends highly on the realized history of signals that he gets.}
In what follows I explain the details of how this example was
implemented in the lab.

@@

I recruited XXX undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
XXX hours and subjects earned an average payment of XXX. The experiment
was programmed using oTree \citep{chen2016otree}.

The experiment consisted of 2 treatments: the \emph{ego-relevant}
condition and the \emph{stereotype} condition. Subjects participated in
only one of the treatments. Treatments were randomly assigned at the
session level. The tasks were identical across treatments, except for
parameter \(\theta\). In the ego-relevant condition \(\theta\) is the
subject's own performance in a quiz, while in the stereotype condition,
it is the performance of a randomly selected subject from another
session.

@@

At the end of the experiment, one of the 6 topics was randomly selected
to determine the payment. They earned \(\$0.20\) for each correct answer
in the quiz, and for each success in part 3.

Randomness is controlled throughout the experiment and sessions by
setting a seed at the beggining of the first session. The seed was drawn
at random and remained fixed for all sessions. By doing this I ensure
that any two subjects who have the same type and face the same exogenous
rate will observe the same outcomes and thus, if they use the same
updating procedure, they should be choosing the same gambles. This
design feature allows me to identify differences in updating procedures
across subjects.

\hypertarget{analysis}{%
\section{Analysis}\label{analysis}}

\renewcommand\refname{Conclusion}
  \bibliography{references.bib}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Learning with misspecified models: Overconfidence and Stereotypes},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\title{Learning with misspecified models: Overconfidence and
Stereotypes}
\author{Jimena Galindo}
\date{October 07, 2023}

@@

\hypertarget{related-literature}{%
\subsection{Related literature}\label{related-literature}}

\hypertarget{theoretical-framework}{%
\section{Theoretical Framework}\label{theoretical-framework}}

I consider multiple theories of belief updating that have been proposed
in the literature and that are able to rationalize the prevalence of
misspecified beliefs. I focus on two different frameworks and develop a
unifying example that allows me to compare the predictions of all the
theories. In this section, I outline the theories within each of the
frameworks.

\hypertarget{framework-1}{%
\subsection{Framework 1}\label{framework-1}}

An agent is of type \(\theta \in \Theta\) and faces an unknown exogenous
state \(\omega\) drawn from some density \(f\) over \(\Omega\). The
agent knows the distribution of \(\omega\) but not its realized value.
His prior belief about his type is \(p_0(\theta)\) and his belief about
the state, \(p_0(\omega)\), coincides with the true distribution. Let
the agent's true type be \(\theta^{*}\), and the realized state be
\(\omega^{*}\).

An agent has a \emph{misspecified} belief if the prior assigns
probability zero to their true type. Furthermore, the agent is
\emph{dogmatic} if he holds a degenerate belief that places probability
one on being a particular type, \(\hat{\theta}\). An agent can be
dogmatic and misspecified. If the agent is both,
\(\hat{\theta} \neq \theta^*\) and \(p_0(\hat{\theta}) = 1\).

@@

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption simplifies the
analysis and plays a role in whether an agent who updates their beliefs
using Bayes rule would learn the truth or not. The behavior of a
forward-looking agent is discussed in the conclusion.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
@Esponda2016 to this setting with only one agent}. A
\emph{self-defeating equilibrium} is a belief and action pair such that
the agent's belief about their type is misspecified and the outcome
generated by the action is consistent with the misspecified belief. This
means that the average outcome under the true type and the true state
equals the average output the agent expects under the misspecified
belief. The agent's belief is said to be \emph{stable} when this
happens.

Within this framework, I consider two nested theories of belief
updating: the first one is a dogmatic modeler from \citet{Heidhues2018},
The second one is a switcher, as in \citet{Ba2023}. The dogmatic modeler
can be seen as a switcher with an infinitely sticky initial belief and
this is the sense in which the two are nested. Both theories produce
different predictions about the agent's behavior. I discuss both in what
follows.

\hypertarget{the-dogmatic-modeler}{%
\subsubsection{The Dogmatic Modeler}\label{the-dogmatic-modeler}}

A dogmatic agent does not update their beliefs about \(\theta\),
instead, he holds a degenerate belief that places probability on
\(\hat{\theta}\), which is potentially misspecified. In this case, no
matter how much information he gathers against being of type
\(\hat{\theta}\), he will not update his beliefs about \(\theta\). Any
discrepancies between the observed outcomes and his believed type are
incorporated using the Bayes rule to update their beliefs about
\(\omega\). \citet{Heidhues2018} show that, under certain assumptions on
the per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{ee}<0$ and $u_e(\underline{e} \theta, \omega)>0>u_e(\Bar{e}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{e\theta}<0$ and $u_{e\omega}>0$. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes they observe reinforce their
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, they will be in a
self-defeating equilibrium with a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome. He expected a
gain of \(h(\hat{\theta}, \mathbb{E}(\omega), a)\) but instead observes
\(h(\hat{\theta}, \omega^*, a)\). There are two possible sources for the
disappointment, the first is that the realized state is lower than the
expected state. The second source is that the agent is of type
\(\theta^*\) and therefore, for all possible states, his gain will be
lower than what he expected. Because the agent is dogmatic, he will not
update his beliefs about \(\theta\) and as a consequence will attribute
the disappointment to the state being lower than expected. He will
continue to update in this way until he converges to a belief about
\(\omega\) that is stable. Such a belief will explain the observed
utility perfectly and allow the agent to rationalize his dogmatic belief
about \(\theta\). Under the assumptions of \citet{Heidhues2018}, there
is a unique value of \(\omega\) at which the belief is stable, I will
refer to such value as \(\omega_\infty\). This mechanism is further
illustrated in Example 1.

@@

\emph{Under this specification, the myopic optimal effort level is
\(a_t^* = \omega^*\). Nonetheless, because the agent does not know
\(\omega^*\), he will choose \(a_t = \mathbb{E}_t(\omega)\) where the
expectation is taken with respect to the agent's belief at the beginning
of period \(t\). If he does not revise his effort choice for \(k\)
periods, he will receive an average utility of
\((\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\) but he was expecting
an average utility of
\((\hat{\theta}\theta+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\). In
response, he will apply Bayes rule to update his beliefs about
\(\omega\) to get the posterior belief with
\(\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}\)
which is lower than the initial belief. This will cause the agent to
choose a lower effort at \(t+k\). As a result, he will again receive an
average utility that is lower than what he expected which will cause his
belief to drift further down. This process will continue until the
average utility equals his expected utility under the dogmatic belief
that assigns probability 1 to \(\hat{\theta}\). At that point, the
student will have reached a self-defeating equilibrium and he will
continue to choose sub-optimal effort forever. }

Although the model of a dogmatic modeler rationalizes the prevalence of
overconfident (underconfident) beliefs, the assumption that the agent
has a degenerate belief and no mechanism through which he can update
such belief is very restrictive. An alternative approach is proposed by
\citet{Ba2023}. She proposes an extension of the dogmatic agent who is
able to jump from one dogmatic belief to another. By doing so, the agent
might end up being dogmatic and correctly specified.

\hypertarget{the-switcher}{%
\subsubsection{The Switcher}\label{the-switcher}}

An agent is a \emph{switcher} if they behave as a dogmatic, but is
willing to entertain the possibility that they are of a different type.
In particular, when they start off as a misspecified dogmatic, they are
willing to switch to a different dogmatic belief if the data is
convincing enough. Their prior is still degenerate and assigns
probability one to a particular type, and zero to all other types. This
means that a Bayesian update on \(\theta\) does not change their beliefs
about the type. However, they are willing to entertain two such beliefs
and have a mechanism by which they decide which belief to adopt at any
period \(t\).

In order to abandon their initial dogmatic belief, the agent needs to
observe a sequence of outcomes that are sufficiently unlikely to have
happened if they were of the type they initially believed. In order to
evaluate if the evidence is convincing enough, they keep track of the
likelihood that each of the possible types generated the data. If the
likelihood ratio is sufficiently large, the agent will switch to the
alternative and behave as if they are dogmatic about the new type.

In particular, for an agent that starts with a dogmatic belief that they
are of type \(\hat{\theta}\) but is willing to consider the alternative
explanation that they are of type \(\tilde{\theta}\), the agent will
switch to the alternative if:

@@

Where \(h^t\) is the history of outcomes up to time \(t\) and \(\alpha\)
is the switching threshold. By keeping track of the likelihood ratio,
the agent can perform a \emph{Bayesian hypothesis test} and adopt the
Dogmatic belief that best fits the
data.\footnote{In a related problem @Schwarstein2021 proposes a similar updating procedure which relies on the Bayesian 
hypothesis test. However, in their model there is a sender who optimally chooses to propose a model that fits the data}
Notice that if \(\alpha \to \infty\), the behavior of the switcher will
be indistinguishable from that of the Dogmatic modeler. In this sense,
the switcher is a generalization of the dogmatic type.

By allowing the agent to keep track of the likelihoods and switching to
an alternative type, the switcher can avoid the self-confirming
equilibrium. However, if the prior belief on \(omega\) is sufficiently
tight around a self-defeating equilibrium, the switcher might look
identical to the dogmatic even in a case where \(\alpha\) is not too
large. This happens because under the agent's prior, the likelihood
ratio is unlikely to grow as fast as it is needed to escape the
self-defeating equilibrium. In such situations, we say that the
misspecified belief is persistent.

\hypertarget{framework-2}{%
\subsection{Framework 2}\label{framework-2}}

As in framework 1, the agent is of some type \(\theta^* \in \Theta\) and
the state is \(\omega \sim F(\Omega)\). In this case, the agent chooses
an action \(a\in A\) and observes a binary outcome that is either a
success or a failure. Denote the outcome by \(o \in {s,f}\). The
probability of observing a success is increasing in \(\theta^*\) and in
\(\omega\). Whenever the agent observes a success, he gets a payoff
\(v>0\) and whenever the outcome is a failure, the payoff is 0. In
addition, the probability of success is such that for each state, there
is a unique optimal action that maximizes the agent's expected payoff.
Therefore, the probability of success can be seen as a monotone
transformation of the utility from Framework 1.

I focus on two nested theories that have been widely studied within this
framework: Full Bayesian updating and self-serving attribution bias. I
explain each of these classical models of belief updating in what
follows

\hypertarget{the-bayesian}{%
\subsubsection{The Bayesian}\label{the-bayesian}}

A Bayesian agent simultaneously updates their beliefs about \(\theta\)
and \(\omega\) by using Bayes' rule. The posterior at period t+1 about
\(\theta\) after observing a history of outcomes \(h^t\) is given by: \[
p_{t+1}(\theta, \omega| h^t) = \frac{p[h^t|\theta, \omega]p_t(\theta, \omega)}{\sum_{(\theta', \omega')}p[h^t|\theta', \omega']p_t(\theta', \omega')}
\]

Where \(p_{t}\) is the belief at the start of period \(t+1\). The update
is symmetric for \(\omega\).

Bayesian agents choose the effort level that maximizes their expected
flow payoff by taking expectations over their prior beliefs about
\(\theta\) and \(\omega\). Since agents are myopic, even though all the
parameters could be identified with enough variation in choices.
Although the Bayesian agent is the closest to a fully rational agent
discussed here, they might not learn their true type. This happens
because, by being myopic, they do not internalize the tradeoff between
flow payoff and learning. This can result in too little experimentation
to learn their true type. An alternative to this approach is given by
\citet{Hestermann2021} and is discussed in the concluding remarks.

Also notice that if a fully Bayesian agent has a dogmatic prior, they
will never update their beliefs about the parameter that they are
dogmatic about. This will imply that they are prone to the same types of
errors as the dogmatic modeler in framework 1.

\hypertarget{the-self-serving-updater}{%
\subsubsection{The Self-Serving
Updater}\label{the-self-serving-updater}}

A self-serving Bayesian is an agent who uses a biased version of Bayes
rule to update their beliefs. They will update their beliefs about the
state \(\omega\) and his type \(\theta\) simultaneously by
over-attributing successes to a high value of \(\theta\) and
under-estimating the role of higher \(\omega\). Similarly, he will
attribute failure to a low state to a greater degree than an unbiased
agent would. To model the self-serving attribution bias, I take the
approach of \citet{benjamin2019}, where the posterior odds are given by:

\[
p_{t+1}(\theta, \omega| h^t) = \frac{p[h^t|\theta, \omega]^{c(\theta, \omega, o_t)}p_t(\theta, \omega)}{\sum_{(\theta', \omega')}p[h^t|\theta', \omega']^{c(\theta, \omega, o_t)}p_t(\theta', \omega')}
\]

with
\(c(\theta_H, \omega, o)<c(\theta_M, \omega, o)<c(\theta_L, \omega, o)\leq 1\)
and
\(c(\theta, \omega_L, o)<c(\theta, \omega_M, o)<c(\theta, \omega_H, o)\leq 1\).

This formulation of the bias means that the agent will over-attribute
successes to a high type and under-attribute them to a low type. Whereas
they will over-attribute failures to a low state and under-attribute
them to a high state. The parameter \(c\) determines the degree of bias
and the restriction on the order ensures that the model is still
falsifiable.

@@

In order to compare the predictions of the theories discussed above, I
develop a unifying example that allows me to isolate the forces behind
each of the theories. The example is a modification of the one in
\citet{Heidhues2018} and is adapeted to be easier to implement in the
lab.

@@

By including self-confirming equilibria, the example captures the forces
from each of the updating mechanisms discussed in the previous section
and allows for the comparison of the main forces behind the theories.
For realizations of \((\theta, \omega)\) for which there are
self-confirming equilibria, the dogmatic agent will fall into the trap
whereas the switcher will be able to escape it. Similarly, an agent with
self-attribution bias will update their beliefs differently from an
unbiased Bayesian, leading them to choose different gambles. I exploit
such cases in order to test which model is a better fit for how subjects
behave in a laboratory experiment.
\footnote{because the setting does not match that of Heidhues et al. [2018], there will be situationf for which the theory 
does not provide a prediction. If such cases arise in the lab, they will not be used for the analysis. However, whether a misspecified 
belief persists or not for the switcher, depends highly on the realized history of signals that he gets.}
In what follows I explain the details of how this example was
implemented in the lab.

@@

I recruited undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{chen2016otree}.

The experiment consisted of 2 treatments: the \emph{ego-relevant}
condition and the \emph{stereotype} condition. Subjects participated in
only one of the treatments. All subjects within a session participated
in the same treatment and the first 4 treatments were assigned the
ego-relevant condition; the rest were assigned to the stereotype
treatment. The tasks were identical across treatments, except for
parameter \(\theta\). In the ego-relevant condition \(\theta\) is the
subject's own performance in a quiz, while in the stereotype condition,
it is the performance of a randomly selected subject from another
session.

@@

At the end of the experiment, one of the 6 topics was randomly selected
to determine the payment. They earned \(\$0.20\) for each correct answer
in the quiz, and for each success in the task in Part 3 for the selected
topic.

Randomness is controlled throughout the experiment and sessions by
setting a seed at the beginning of the first session. The seed was drawn
at random and remained fixed for all sessions.
\footnote{The seed that was drawn 
at the beginning of session 1 was 3452. The same seed was used for all sessions. It is used both for drawing $\omega$ for each of 
the tasks in the experiment, as well as for drawing the outcomes from the gambles.}
By doing this I ensure that any two subjects who have the same type and
face the same exogenous rate will observe the same outcomes and thus, if
they use the same updating procedure, they should be choosing the same
gambles. This design feature allows me to identify differences in
updating procedures across subjects.
